---
title: "EDA"
author: "Joshua Tolhuis"
date: "9/14/2021"
output:
  pdf_document: default
  html_document: default
---

## Assignment Introduction:
In this assignment I was tasked to research an existing data file given to me and apply data mining & machine learning to predict the outcome. I first was tasked creating a research question around the data. My question went as follows:
"What's the minimal amount of data combinations to predict a patient has pancreatic ductal adenocarcinoma?"


## Data

The data gained was from John Davis, uploaded to kaggle [John Davis](https://www.kaggle.com/johnjdavisiv/urinary-biomarkers-for-pancreatic-cancer).
He had gotten his data from a paper with the title "A combination of urinary biomarker panel and PancRisk score for earlier detection of pancreatic caner: A case-control study", this is the [link](https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.1003489) to the paper.  

setup and load code book:  

Loading up libraries to be used in the EDA and log.  
```{r, results='hide', message=FALSE, warning=FALSE}
if(!require(devtools)) install.packages("devtools")
devtools::install_github("sinhrks/ggfortify")
devtools::install_github("AckerDWM/gg3D")
library("gg3D")
library(ggplot2)
library(tidyr)
library(gridExtra) 
library(dplyr)
library(plotly)
library(cluster)
library(ggfortify)
```

Let's begin with loading in the required data and the code book describing it.    
```{r}
setwd("./Data")

data <- read.csv("Debernardi et al 2020 data.csv", header = T, sep = ",")
str(data[])

codebook <- read.csv("Debernardi et al 2020 documentation.csv", sep = ",", header = T)

knitr::kable(codebook[2:3])
```

## Intro
At first, I wanted to know if REG1B could be used only to predict PDAC, but after reconsideration I want to find out what the minimum required data is to find out if a patient has PDAC. Even so i first wanted to run some tests on the newly improved biomarker REG1B to find out it's impact, and it's change relative to REG1A.  


## Results
Let's take a first look at the effect of REG1B stand alone, what conclusions can be made by looking at the markers found at different stages.  
```{r}
ggplot(data = data, mapping = aes(x = stage, y = REG1B, col = stage)) +
        geom_point(alpha = 0.7) +
        geom_jitter() +
        xlab("Stages of PDAC (I-IV)") +
        ylab("REG1B (ng/mL)") +
        ggtitle("Stages of PDAC versus REG1B")
```

In this plot becomes clear, that REG1B has the same values at the stages "0, IIB, III and IV" relative to the others. This immediately shuts down the proposition to predict stages of cancer with only 1 variable.  

In order to find out more I decided to view the other bio markers and this time with a box plot to see if there was a recognizable trend.I tried several box plots on the bio markers, first with normal data, second with scaled data, third with outliers removed and the last plot has been scaled and the outliers removed from it. I also had to create functions for the latter 3 plots.  

scaling function:  
```{r}
scale_min_max <- function(x) {
     (x - min(x)) / (max(x) - min(x))
}
```

outlier removal function:  
```{r}
remove_outliers <- function(x, na.rm = TRUE, ...) {
  qnt <- quantile(x, probs=c(.25, .75), na.rm = na.rm, ...)
  H <- 1.5 * IQR(x, na.rm = na.rm)
  y <- x
  y[x < (qnt[1] - H)] <- NA
  y[x > (qnt[2] + H)] <- NA
  y
}
```

I printed them all separately to get a better view.  
```{r}

scaled_data <- data

scaled_data$REG1B <- scale_min_max(scaled_data$REG1B)
scaled_data$TFF1 <- scale_min_max(scaled_data$TFF1)
scaled_data$LYVE1 <- scale_min_max(scaled_data$LYVE1)

p1 <- pivot_longer(data = data, cols = c(REG1B,TFF1,LYVE1), names_to = "Var", values_to = "Val") %>%
        ggplot(aes(x = Var, y = Val, fill = stage)) +
        geom_boxplot() +
        xlab("Stages of PDAC (I-IV)") +
        ylab("Biomarkers (ng/mL)") +
        ggtitle("Stages of PDAC versus biomarkers")

p2 <- pivot_longer(data = scaled_data, cols = c(REG1B,TFF1,LYVE1), names_to = "Var", values_to = "Val") %>%
        ggplot(aes(x = Var, y = Val, fill = stage)) +
        geom_boxplot() +
        xlab("Stages of PDAC (I-IV)") +
        ylab("Biomarkers (scaled 0 - 1)") +
        ggtitle("Stages of PDAC versus biomarkers")

p3<- pivot_longer(data = data, cols = c(REG1B,TFF1,LYVE1), names_to = "Var", values_to = "Val") %>%
        ggplot(aes(x = Var, y = remove_outliers(Val), fill = stage)) +
        geom_boxplot()

p4 <- pivot_longer(data = scaled_data, cols = c(REG1B,TFF1,LYVE1), names_to = "Var", values_to = "Val") %>%
        ggplot(aes(x = Var, y = remove_outliers(Val), fill = stage)) +
        geom_boxplot() +
        xlab("Stages of PDAC (I-IV)") +
        ylab("Biomarkers (scaled 0 - 1)") +
        ggtitle("Stages of PDAC versus biomarkers")

p1 #normal data
p2 #scaled data
p3 #outliers removed
p4 #scaled data and outliers removed

```

From these plots a few things are noticeable. First of all the without outliers the difference between stage 0 and other stages becomes compelling. This is good to know to be able to make predictions, but seeing as this is cancer and false positives or false negatives are dangerous. outliers will be necessary to make valid predictions. the second noticeable thing is that there is indeed a trend within the bio markers and stages. This is precious information to be able to predict the stages of cancer.  

To find out the most prominent factor within the bio markers i decided to plot them again individually  
```{r}
p1 <- ggplot(data = data, mapping = aes(x = stage, y = scale_min_max(LYVE1), fill = stage)) +
        geom_boxplot(alpha = 0.7) +
        xlab("Stages of PDAC (I-IV)") +
        ylab("Biomarkers (scaled 0 - 1)")

p2 <- ggplot(data = data, mapping = aes(x = stage, y = scale_min_max(TFF1), fill = stage)) +
        geom_boxplot(alpha = 0.7) +
        xlab("Stages of PDAC (I-IV)") +
        ylab("Biomarkers (scaled 0 - 1)")

p3 <- ggplot(data = data, mapping = aes(x = stage, y = scale_min_max(REG1B), fill = stage)) +
        geom_boxplot(alpha = 0.7) +
        xlab("Stages of PDAC (I-IV)") +
        ylab("Biomarkers (scaled 0 - 1)")

grid.arrange(p1, p2, p3, nrow = 2)


```


After this I wanted to know if age might be an important factor to the output of the bio markers so i tried a plot and added a trend line. 

```{r}
p1 <- ggplot(data = data, mapping = aes(x = age, y = LYVE1)) +
        geom_point(alpha = 0.7) + 
        geom_smooth(method = "loess", formula = "y ~ x") +
        xlab("Age") +
        ylab("LYVE1 (ng/mL)") +
        ggtitle("Age against LYVE1")

p2 <- ggplot(data = data, mapping = aes(x = age, y = TFF1)) +
        geom_point(alpha = 0.7) + 
        geom_smooth(method = "loess", formula = "y ~ x") +
        xlab("Age") +
        ylab("TFF1 (ng/mL)") +
        ggtitle("Age against TFF1")

p3 <- ggplot(data = data, mapping = aes(x = age, y = REG1B)) +
        geom_point(alpha = 0.7) + 
        geom_smooth(method = "loess", formula = "y ~ x") +
        xlab("Age") +
        ylab("REG1B (ng/mL)") +
        ggtitle("Age against REG1B")
grid.arrange(p1, p2, p3, nrow = 2)
```

In these plots I concluded that age indeed has an effect on the outcome of the 'LYVE1' bio marker. The other 2 also had a ascending trend line but they show less of an increase that LYVE1.  

There were also creatinine, and blood plasma factors within the data.  
In order to find more correlations I decided to create a heat map to find more correlation within these variables.  
```{r}

selection <- c("age", "plasma_CA19_9", "creatinine", "LYVE1", "TFF1", "REG1B", "REG1A")

tmp <- data %>% select(selection) %>% drop_na()

cor_matrix <- cor(tmp)

cor_matrix <- as_tibble(cor_matrix)
(cor_matrix <- cor_matrix %>% mutate(var1=selection) %>% select(8,1:7))

cor_matrix_long <- pivot_longer(data = cor_matrix, cols = selection, names_to = "variable", values_to = "cor")


ggplot(data = cor_matrix_long, aes(x=var1, y=variable, fill=cor)) + 
    geom_tile() + 
    labs(x=NULL, y=NULL) + 
    scale_fill_gradient(high = "red", low = "white" ) +
        theme(axis.text.x = element_text(angle = 90)) +
        ggtitle("Correlation between numerical values")
```
This heat map shows that bio markers have a high correlation, TFF1 and LYVE1 look like they have the highest correlation.
LYVE1 is indeed as predicted very correlated to age and TFF1 and LYVE1 to creatinine. It's also clear that REG1A has been improved considerably and REG1B has a much higher correlation to other factors.

In order to further look at the correlation between the bio markers I decided to create a PCA.  
```{r}
df <- data[c(11:13)]
  #clustered by stage
pca_res <- prcomp(df, scale = TRUE)
  autoplot(pca_res, data = data, colour = 'stage', frame = TRUE, frame.type = 'norm')
  
  
  #clustered by bio marker
  autoplot(pam(data[11:13], 3), frame = TRUE, frame.type = 'norm')
```

Here I visualized the clusters decided by stage, and the second plot I clustered by bio marker.  

## Conclusion

I concluded that it should be possible to decide the stage of cancer by less data than given. The 3 bio markers should give a clear prediction and this can be helped with the age factor.

## Criteria 
For this research i Think the most valuable is accuracy, speed isn't necessary as I don't think it matters if takes 1 or 10 seconds to process. online classification also isn't important because o normal person probably doesn't have data on bio markers in his presence. Accuracy is the most important as False negatives or False positives are extremely dangerous.

## Cleaning data
To use the data in an algorithm, a few columns should be removed that aren't useful.
first of all the benign sample data, patient cohort and sample origin. I will also remove REG1A as it has a lot of NA values and REG1B is a better version of this bio marker. I will also remove blood plasma as there are again a lot of NA values and the goal of this research is using the least amount of data_values anyway. I also made stage "" into stage 0 as weka doesn't understand "".

```{r}

data$stage <- replace(data$stage,0:391, "none")

row.has.na <- apply(data[1:13], 1, function(x){any(is.na(x))})
filtered <- data[!row.has.na,]

data$cancer <- data[6]

data$


write.csv(filtered[c(4:7 ,9:13)], "Cleaned_data.csv")
```

## Results WEKA
in the WEKA I was able to use the logistics algorithm to gain a 75% correct model, and with the stages added to the list i was able to get it to 86%. Adding the stages however is like cheating, as of course you could predict if somebody has malignant cancer if you know he's in the 3rd or 4th stage. Still I saved the model as the research paper where the data was gained from achieved a >80% SP and SN rate while using this. I was able to get the model to 75% without the use of age and sex. But it seemed that the other variables only decreased the percentage of correct predictions if removed.

## ROC
Here we are looking at the ROC.
```{r}
roc_data <- read.table("Data/ROC.arff", 
                   sep = ",", 
                   comment.char = "@")
names(roc_data) <- c("Instance_number", 
                 "True_Positives",
                 "False_Negatives",
                 "False_Positives",
                 "True_Negatives",
                 "False_Positive_Rate",
                 "True_Positive_Rate",
                 "Precision",
                 "Recall",
                 "Fallout",
                 "FMeasure",
                 "Sample_Size",
                 "Lift",
                 "Threshold")

 

head(roc_data)

 

library(ggpubr)
colors <- c(classifier = "red", threshold = "blue")
plt1 <- ggplot(data = roc_data,
       mapping = aes(x = False_Positive_Rate, y = True_Positive_Rate)) +
    geom_point(mapping = aes(color = "classifier")) +
    geom_abline(aes(color = "threshold", 
                    slope = 1, 
                    intercept = 0)) + 
    scale_color_manual(values = colors) +
    xlab("True Positive Rate") +
    ylab("False Positive Rate") +
    #theme_minimal() +
    theme_pubr() +
    theme_minimal() +
    theme(legend.title = element_blank())


roc_data <- read.table("Data/ROC2.arff", 
                   sep = ",", 
                   comment.char = "@")
names(roc_data) <- c("Instance_number", 
                 "True_Positives",
                 "False_Negatives",
                 "False_Positives",
                 "True_Negatives",
                 "False_Positive_Rate",
                 "True_Positive_Rate",
                 "Precision",
                 "Recall",
                 "Fallout",
                 "FMeasure",
                 "Sample_Size",
                 "Lift",
                 "Threshold")

 

head(roc_data)

 

library(ggpubr)
colors <- c(classifier = "red", threshold = "blue")
plt2 <- ggplot(data = roc_data,
       mapping = aes(x = False_Positive_Rate, y = True_Positive_Rate)) +
    geom_point(mapping = aes(color = "classifier")) +
    geom_abline(aes(color = "threshold", 
                    slope = 1, 
                    intercept = 0)) + 
    scale_color_manual(values = colors) +
    xlab("True Positive Rate") +
    ylab("False Positive Rate") +
    #theme_minimal() +
    theme_pubr() +
    theme_minimal() +
    theme(legend.title = element_blank())


roc_data <- read.table("Data/ROC3.arff", 
                   sep = ",", 
                   comment.char = "@")
names(roc_data) <- c("Instance_number", 
                 "True_Positives",
                 "False_Negatives",
                 "False_Positives",
                 "True_Negatives",
                 "False_Positive_Rate",
                 "True_Positive_Rate",
                 "Precision",
                 "Recall",
                 "Fallout",
                 "FMeasure",
                 "Sample_Size",
                 "Lift",
                 "Threshold")

 

head(roc_data)

 

library(ggpubr)
colors <- c(classifier = "red", threshold = "blue")
plt3 <- ggplot(data = roc_data,
       mapping = aes(x = False_Positive_Rate, y = True_Positive_Rate)) +
    geom_point(mapping = aes(color = "classifier")) +
    geom_abline(aes(color = "threshold", 
                    slope = 1, 
                    intercept = 0)) + 
    scale_color_manual(values = colors) +
    xlab("True Positive Rate") +
    ylab("False Positive Rate") +
    #theme_minimal() +
    theme_pubr() +
    theme_minimal() +
    theme(legend.title = element_blank())


print(plt1)
print(plt2)
print(plt3)
```