---
title: "Log"
author: "Joshua Tolhuis"
date: "9/14/2021"
output: html_document
---

# EDA of Urinary Biomarkers for cancer

## Assignment Introduction:

 *insert text here*
 
## Data

The data gained was from John Davis, uploaded to kaggle [John Davis](https://www.kaggle.com/johnjdavisiv/urinary-biomarkers-for-pancreatic-cancer).
He had gotten his data from a paper with the title "A combination of urinary biomarker panel and PancRisk score for earlier detection of pancreatic caner: A case-control study", this is the [link](https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.1003489) to the paper.

setup and load code book:
```{r}
codebook <- read.csv("Debernardi et al 2020 documentation.csv", sep = ",", header = T)

knitr::kable(codebook)
```

Here is the corresponding data file and it's first 6 entries:
```{r}
data <- read.csv("Debernardi et al 2020 data.csv", header = T, sep = ",")
knitr::kable(head(data))
```

Loading up libraries to be used in the EDA and log
```{r}
if(!require(devtools)) install.packages("devtools")
devtools::install_github("sinhrks/ggfortify")
devtools::install_github("AckerDWM/gg3D")
library("gg3D")
library(ggplot2)
library(tidyr)
library(gridExtra) 
library(dplyr)
library(plotly)
library(cluster)
library(ggfortify)

```

## Intro
At first, I wanted to know if REG1B could be used only to predict PDAC, but after reconsideration I want to find out what the minimum required data is to find out if a patient has PDAC. Even so i first wanted to run some tests on the newly improved biomarker REG1B to find out it's impact, and it's change relative to REG1A.

Let's take a first look at the effect of REG1B stand alone, what conclusions can be made by looking at the markers found at different stages.
```{r}
ggplot(data = data, mapping = aes(x = stage, y = REG1B)) +
        geom_point(alpha = 0.5) +
        xlab("Stages of PDAC (I-IV)") +
        ylab("REG1B (ng/mL)")
```
In this plot becomes clear, that REG1B is has teh same values at the stages "0, IIB, III and IV" relative to the others. This rises the question with which other data values the results be improved so that stage 0 and I - IIA can be recognized more easily. Lets first try finding patterns using only bodily data such as, age, sex, and body fluids or hormones.

```{r}
age.groups <- cut(data$age, breaks = c(25 ,40, 60, 80, 90))
ggplot(data = data, mapping = aes(x = stage, y = LYVE1, col = age.groups)) +
        geom_point(alpha = 0.5) +
        geom_jitter()
        
```

in this plot the REG1B has gained a third dimension, this dimension is age. I wanted to figure out if age might have an influence on the output of the bio marker. it looks like this however is not the case, age is pretty randomly diveded except for people under 40. 

```{r}
sum(data$age < 40)
sum(data$age < 40 & data$stage == "")
```

But I found out that there were only 43 people under 40 in this data set and only 3 of them had a state assigned to them. So I concluded, age had no effect on the effects of this biomarker.

I also tried another few values as 3rd dimension to see if any would be promising:
```{r}

creatinine.groups <- cut(data$creatinine, breaks = c(0.05654 ,1.071623 ,2.086695, 3.101768, 4.11685))
ggplot(data = data, mapping = aes(x = stage, y = creatinine)) +
        geom_boxplot() +
        geom_jitter(alpha = 0.7, aes(col = data$LYVE1))

```

```{r}
data %>% select(REG1B, TFF1, LYVE1, stage) %>%
pivot_longer(., cols = c(REG1B,TFF1,LYVE1), names_to = "Var", values_to = "Val") %>%
        ggplot(aes(x = Var, y = Val, fill = stage)) +
        geom_boxplot()
        



```


```{r}
p1 <- ggplot(data = data, mapping = aes(x = stage, y = LYVE1, fill = stage)) +
        geom_boxplot(alpha = 0.7) +
        ylim(0, 15)

p2 <- ggplot(data = data, mapping = aes(x = stage, y = TFF1, fill = stage)) +
        geom_boxplot(alpha = 0.7) +
        ylim(0, 5000)

p3 <- ggplot(data = data, mapping = aes(x = stage, y = REG1B, fill = stage)) +
        geom_boxplot(alpha = 0.7)

grid.arrange(p1, p2, p3, nrow = 2)


```

```{r}
scale_min_max <- function(x) {
     (x - min(x)) / (max(x) - min(x))
}

df <- as.data.frame(data[392:590,])
row.names(df) <- paste(df$stage, row.names(df), sep="_")

df[11:13]
autoplot(pam(df[c(8,11:13)], 3), frame = TRUE, frame.type = 'norm')


ggplot(df, aes(x= scale_min_max(TFF1), y = scale_min_max(LYVE1), z = scale_min_max(REG1B), color= stage)) + 
  theme_void() +
  axes_3D() +
  stat_3D() +
        xlab("TFF1")

```